Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-07-28 writeup/conStruct.tex
added xvalidation section

\gc{ 
\paragraph{Choice of cluster number and cross-validation}

@@

In contrast, the predictive accuracy of the spatial model of \texttt{conStruct} 
either plateaus at or declines from the true value of $K$ used to simulate the data (Fig \ref{xvals}, second column).
And, as $K$ in the spatial model increases, 
deme membership in clusters beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.

@@

\subsection*{Empirical Applications}

@@

@@

\newpage
\section*{Appendix}
\renewcommand{\theequation}{A\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

@@

\newpage
\section{Cross validation}
To perform model comparison, we employ a Monte Carlo cross-validation approach \citep{picard1984},
also known as repeated random sub-sampling validation.
Briefly, we follow the following procedure:
\begin{enumerate}
\item For each of $X$ replicates:
	\begin{enumerate}
		\item partition the allele frequency data into a 90\% ``training'' partition ($F^x_1$) and a 10 \%``testing" partition ($F^x_2$) \label{partition}
		\item run our inference procedure on the training partition to estimate model parameters $\theta_{mk}$ for: \label{inference}
			\begin{enumerate}
				\item $m$: the spatial and the nonspatial model
				\item $k$: the number of clusters 1 through $K$
			\end{enumerate}	
	\item calculate the mean log likelihood of the testing data partition, 
	over the posterior distribution of training-estimated parameters for each model 
	($\bar{\mathcal{L}}(F^x_2 \mid \theta_{mk})$, henceforth $\bar{\mathcal{L}}_{xmk}$) \label{lnL}
	\item generate standardized mean log likelihoods, $\mathcal{Z}_{xmk}$, across models: \label{standardize}
		\begin{enumerate}
			\item identify the highest mean log likelihood, $\bar{\mathcal{L}}^\text{max}_{xmk}$
			\item subtract $\bar{\mathcal{L}}^\text{max}_{xmk}$ from $\bar{\mathcal{L}}_{xmk}$ across all models,
				such that the standardized log likelihood, $\mathcal{Z}_{xmk}$, of the best model is 0,
				and less than 0 for all inferior models. 
		\end{enumerate}
	\end{enumerate}
\item For each model (i.e., each combination of $m$ and $k$) calculate 
	the mean standardized log likelihood of the testing data partition across $X$ replicates, 
	as well its standard error and 95\% confidence interval:
	\begin{enumerate}
	\item mean
		\begin{equation}
			\bar{\mathcal{Z}}_{mk} = \frac{1}{X}\sum\limits_{x=1}^{X}\mathcal{Z}_{xmk}
		\end{equation}
	\item standard error
		\begin{equation}
			SE_{\bar{\mathcal{Z}}_{mk}} = \sqrt{\frac{\sum\limits_{x=1}^{X} \left[ {\mathcal{Z}_{xmk} - \bar{\mathcal{Z}}_{mk}} \right]}{X}}
		\end{equation}
	\item 95\% confidence interval
		\begin{equation}
			95\% \text{CI} = \bar{\mathcal{Z}}_{mk} \pm 1.96 \times SE_{\bar{\mathcal{Z}}_{mk}}
		\end{equation}
	\end{enumerate}
\end{enumerate}

>>>>

\paragraph{Choice of cluster number and cross-validation}
There are a number of reasons, both statistical and biological, 
why choosing a single, ``best" number of clusters to describe the data may not be desirable.
%, including ambiguity in what exactly it means for any given value of $K$ to be ``best."
We discuss these reasons further in the Discussion below, 
but we also recognize that in many circumstances it is either useful or convenient 
to present only a single set of results from a clustering analysis with a particular number of clusters.
ideally a number large enough to describe meaningful partitions in the data, 
but not so large as to be difficult to parse visually.
Toward that end, we have implemented a method for 
statistically comparing \texttt{conStruct} runs across different values of $K$.

Several approaches have been proposed to determine the number of clusters in the data, 
including: 
the use of information theoretic criteria \citep[e.g.][]{ADMIXTURE}, such as AIC or BIC;
comparisons of the likelihood of the data across different values of $K$, 
with various criteria on how to choose a single value (e.g., Evanno 2005);
or comparisons of the marginal likelihood, 
generated either via various approximations \citep[e.g.][]{STRUCTURE}
or via a more rigorous and computationally intensive approach such as thermodynamic integration \citep{verity_nichols2016}
or inference using a Dirichlet process prior \citep{huelsenbeck2007inference}
A comprehensive review of these approaches and their various merits is beyond the scope of this paper,
but see \cite{verity_nichols2016} for an excellent discussion of and comparison between several.

We choose to take a cross-validation approach, 
estimating the parameters of the model using a ``training" partition of the data,
then calculating the likelihood of a separate, ``testing" partition of the data given those parameters,
repeating this procedure many times, 
and selecting the model with the significantly best predictive accuracy
(see Section \ref{Xvalidation} for a thorough discussion of our cross-validation procedure).
This approach rewards models with larger values of $K$, which increase the likelihood of the training partition,
up to the point that cluster membership is invoked to model noise specific to the training data, 
at which point predictive accuracy decreases, 
and models with additional clusters are penalized.

@@

In contrast, the predictive accuracy of the spatial model of \texttt{conStruct} 
either plateaus at or declines from the true value of $K$ used to simulate the data (Fig \ref{xvals}, second column).
And, as $K$ in the spatial model increases, 
deme membership in clusters beyond the number used to simulate the data does not appreciably increase 
(Fig \ref{simK1_sp_pies}, \ref{simK2_sp_pies}, \ref{simK3_sp_pies}).
That is, even when the spatial model is run with $K=7$, 
the inferred admixture proportions are nearly identical to 
those estimated under the true value of $K$ for each simulation.
We parse \texttt{conStruct}'s performance on these simulation tests further in the Discussion.

@@

\subsection*{Empirical Applications}
To demonstrate the utility of this method, we analyzed empirical population genomic data from two systems:
a contact zone between two poplar species in northwestern North America,
and a large North American sample of black bears.

\subsubsection*{Poplars}

\subsubsection*{Black bears}

@@

\paragraph{Performance on simulated data}
\paragraph{Empirical results}
\paragraph{Choosing the ``best" number of clusters}
\paragraph{Implications for management and conservation}
\paragraph{Extensions and future directions}

@@

\newpage
\section*{Appendix}
\renewcommand{\theequation}{A\arabic{equation}}
\setcounter{equation}{0}
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{table}{0}
\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thesection}{A\arabic{section}}
\setcounter{section}{0}

@@

\newpage
\section{Cross validation}\label{Xvalidation}
To perform model comparison, we employ a Monte Carlo cross-validation approach \citep{picard1984},
also known as repeated random sub-sampling validation.
Briefly, we follow the following procedure:
\begin{enumerate}
\item For each of $X$ replicates:
	\begin{enumerate}
		\item partition the allele frequency data into a 90\% ``training'' partition ($F^x_1$) and a 10 \%``testing" partition ($F^x_2$) \label{partition}
		\item run our inference procedure on the training partition to estimate model parameters $\theta_{mk}$ for: \label{inference}
			\begin{enumerate}
				\item $m$: the spatial and the nonspatial model
				\item $k$: the number of clusters 1 through $K$
			\end{enumerate}	
	\item calculate the mean log likelihood of the testing data partition, 
	over the posterior distribution of training-estimated parameters for each model 
	($\bar{\mathcal{L}}(F^x_2 \mid \theta_{mk})$, henceforth $\bar{\mathcal{L}}_{xmk}$) \label{lnL}
	\item generate standardized mean log likelihoods, $\mathcal{Z}_{xmk}$, across models: \label{standardize}
		\begin{enumerate}
			\item identify the highest mean log likelihood, $\bar{\mathcal{L}}^\text{max}_{xmk}$
			\item subtract $\bar{\mathcal{L}}^\text{max}_{xmk}$ from $\bar{\mathcal{L}}_{xmk}$ across all models,
				such that the standardized log likelihood, $\mathcal{Z}_{xmk}$, of the best model is 0,
				and less than 0 for all inferior models. 
		\end{enumerate}
	\end{enumerate}
\item For each model (i.e., each combination of $m$ and $k$) calculate 
	the mean standardized log likelihood of the testing data partition across $X$ replicates, 
	as well its standard error and 95\% confidence interval:
	\begin{enumerate}
	\item mean
		\begin{equation}
			\bar{\mathcal{Z}}_{mk} = \frac{1}{X}\sum\limits_{x=1}^{X}\mathcal{Z}_{xmk}
		\end{equation}
	\item standard error
		\begin{equation}
			SE_{\bar{\mathcal{Z}}_{mk}} = \sqrt{\frac{\sum\limits_{x=1}^{X} \left[ {\mathcal{Z}_{xmk} - \bar{\mathcal{Z}}_{mk}} \right]}{X}}
		\end{equation}
	\item 95\% confidence interval
		\begin{equation}
			95\% \text{CI} = \bar{\mathcal{Z}}_{mk} \pm 1.96 \times SE_{\bar{\mathcal{Z}}_{mk}}
		\end{equation}
	\end{enumerate}
\end{enumerate}
